---
title: "MASLD Clustering"
author: "Eren Kumak"
date: "2025-01-02"
output: 
  html_document:
   toc: true
   toc_float: 
     collapsed: false
     smooth_scroll: true
   number_sections: true
---

# Introduction

The aim of this document to create PCA plots of each dataset presented in meta-analysis to understand how each NAFLD Activity Score clustered. 

Document is organized into sections each representing an individual dataset where name is direct GEO accession ID and only PCA with points according to NAS identitiy presented. 

Since the process is repetitive, first chunk of this markdown contains functions which will be called later. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(org.Hs.eg.db)
library(GEOquery)
library(Rtsne)
library(ggfortify)
library(dplyr)
library(gridExtra)
library(sva)
library(limma)
library(DESeq2)
library(edgeR)
library(plotly)
library(rmarkdown)
library(vegan)
library(pheatmap)
```


```{r functions}

load_dataset <- function(id){
  a <- fread(paste(id, ".tsv", sep="")) #load the count optained from NCBI-generated RNA-seq
  gene_names <- mapIds(
    org.Hs.eg.db,
    keys = as.character(a$GeneID),
    column = "SYMBOL", 
    keytype = "ENTREZID",
    multiVals = "first"
) # Get the gene symbols from ENTREZIDs
  a$GeneID <- gene_names #assign gene names instead of ENTREZID
  colnames(a)[1] <- "Gene" #change the column name
  a <- a[!duplicated(a$Gene),] #remove duplicated gene names
  a <- a[!is.na(a$Gene),] #remove any NAs in gene name
  a <- as.data.frame(a) #turn to df
  rownames(a) <- a$Gene #assign gene names as row names.
  a$Gene <- NULL #remove column
  return(a) #return fully modified df
}

load_meta <- function(id){
  a  <- getGEO(id, GSEMatrix = TRUE) #get the meta data via accession.
  a  <- pData(a[[1]]) #get the df portion
  return(a) #return it
}

pca_nas_2d <- function(data,meta,col_index, name = "RNA-seq"){
  meta <- meta[rownames(meta) %in% colnames(data),] #Be sure you have meta data for only ones that has counts.
  colData <- as.data.frame(meta[,col_index]) # create the design df using the col_index for activation score.
  rownames(colData) <- rownames(meta) #assign sample names as rownames.
  print(sum(rownames(colData) != colnames(data))) # check whether rownames of colData is same with colnames of data !should return 0
  colnames(colData) <- "score" #name NAS as score
  colData$score <- as.factor(colData$score) #factorize them
  
  dds <- DESeqDataSetFromMatrix(countData = data, colData = colData, design = ~ score) 

  dds <- DESeq(dds) 
  group_sizes <- table(colData$score) #Counts how many samples are in each score group.
  smallest_group_size <- min(group_sizes) #Stores the size of the smallest group 
  
  keep <- rowSums(counts(dds) >= 10) >= smallest_group_size #Keeps genes that are expressed (b	%10 reads) in at least as many samples as the smallest group   size. Example: If smallest group has 3 samples, this line keeps genes that are expressed in b	%3 samples.
  dds <- dds[keep,] #Applies that filter to the dataset.
  
 # Apply Variance Stabilizing Transformation (VST)
  vst_counts <- vst(dds, blind = FALSE)  # 'blind = FALSE because we will use score information in below analysis.

  # Extract the transformed count matrix
  vst_matrix <- assay(vst_counts) 

  variances <- apply(vst_matrix, 1, var) #get variances of each gene
  top_genes <- vst_matrix[variances > quantile(variances, 0.75), ] #get genes whose variance is in the top 25%.
  
  pca_result <- prcomp(t(top_genes), scale = TRUE) #performs PCA on the transposed top_genes matrix so that samples are rows and genes are columns.
  
 autoplot(pca_result, data = colData, colour = 'score') +
    theme_minimal() +
    ggtitle(paste("2D PCA of ", name, " colored by NAFLD Activity Score", sep ="")) ##plot the results, color by score.
}

pca_nas_3d <- function(data,meta,col_index, name = "RNA-seq"){
  meta <- meta[rownames(meta) %in% colnames(data),] #Be sure you have meta data for only ones that has counts.
  colData <- as.data.frame(meta[,col_index]) # create the design df
  rownames(colData) <- rownames(meta) #assign sample names as rownames.
  print(sum(rownames(colData) != colnames(data))) # check whether rownames of colData is same with colnames of data !should return 0
  colnames(colData) <- "score" #name NAS as score
  colData$score <- as.factor(colData$score) #factorize them
  
  dds <- DESeqDataSetFromMatrix(countData = data, colData = colData, design = ~ score) 
  dds <- DESeq(dds)
  
  group_sizes <- table(colData$score) #Counts how many samples are in each score group.
  smallest_group_size <- min(group_sizes) #Stores the size of the smallest group 
  
  keep <- rowSums(counts(dds) >= 10) >= smallest_group_size #Keeps genes that are expressed (b	%10 reads) in at least as many samples as the smallest group   size. Example: If smallest group has 3 samples, this line keeps genes that are expressed in b	%3 samples.
  dds <- dds[keep,] #Applies that filter to the dataset.
  

  # Apply Variance Stabilizing Transformation (VST)
  vst_counts <- vst(dds, blind = FALSE)   # 'blind = FALSE because we will use score information in below analysis.

  # Extract the transformed count matrix
  vst_matrix <- assay(vst_counts)

  variances <- apply(vst_matrix, 1, var) #get variances of each gene
  top_genes <- vst_matrix[variances > quantile(variances, 0.75), ] #get genes whose variance is in the top 25%.
  
  pca_result <- prcomp(t(top_genes), scale = TRUE) #performs PCA on the transposed top_genes matrix so that samples are rows and genes are columns.
  pca_data_3d <- as.data.frame(pca_result$x[, 1:3])  # First 3 PCs
  
 plot_ly(pca_data_3d, x = ~PC1, y = ~PC2, z = ~PC3, color = colData$score,
                 colors = RColorBrewer::brewer.pal(9, "Set3"), 
                 type = "scatter3d", mode = "markers") %>%
    layout(title = paste("3D PCA of ", name, " colored by NAFLD Activity Score", sep = ""),
           scene = list(xaxis = list(title = 'PC1'),
                        yaxis = list(title = 'PC2'),
                        zaxis = list(title = 'PC3'))) #plot results interactively.
}

statistical_report_pca <- function(data,meta,col_index, name = "RNA-seq"){
  meta <- meta[rownames(meta) %in% colnames(data),] #Be sure you have meta data for only ones that has counts.
  colData <- as.data.frame(meta[,col_index]) # create the design df
  rownames(colData) <- rownames(meta)  #assign sample names as rownames.
  print(sum(rownames(colData) != colnames(data))) # check whether rownames of colData is same with colnames of data !should return 0
  colnames(colData) <- "score" #name NAS as score
  colData$score <- as.factor(colData$score) #factorize them
  
  dds <- DESeqDataSetFromMatrix(countData = data, colData = colData, design = ~ score) 
  dds <- DESeq(dds)
  
  group_sizes <- table(colData$score) #Counts how many samples are in each score group.
  smallest_group_size <- min(group_sizes) #Stores the size of the smallest group 
  
  keep <- rowSums(counts(dds) >= 10) >= smallest_group_size #Keeps genes that are expressed (b	%10 reads) in at least as many samples as the smallest group   size. Example: If smallest group has 3 samples, this line keeps genes that are expressed in b	%3 samples.
  dds <- dds[keep,] #Applies that filter to the dataset.
  
  
  # Apply Variance Stabilizing Transformation (VST)
  vst_counts <- vst(dds, blind = FALSE)  # 'blind = FALSE because we will use score information in below analysis.

  # Extract the transformed count matrix
  vst_matrix <- assay(vst_counts)

  variances <- apply(vst_matrix, 1, var) #get variances of each gene
  top_genes <- vst_matrix[variances > quantile(variances, 0.75), ] #get genes whose variance is in the top 25%


  # Initialize an empty matrix to store the p-values for pairwise comparisons
  p_value_matrix <- matrix(NA, nrow = length(unique(colData$score)), 
                           ncol = length(unique(colData$score)))

  # Set the group labels as row and column names
  rownames(p_value_matrix) <- colnames(p_value_matrix) <- sort(unique(colData$score))
  
  vst_group <- as.data.frame(t(vst_matrix))
  vst_group$Group <- as.vector(colData$score)
  
  dist_matrix <- dist(vst_group[, -ncol(vst_group)])  # Adjust if you want to use more PCs
    perm_results <- adonis2(dist_matrix ~ vst_group$Group, permutations = 999)
    print("PERMANOVA RESULTS FOR Variance Stabilizing Transformation (VST)")
    cat("\n") 
    print("############################")
    print(perm_results)
    print("############################")
    cat("\n") 
  
  # Perform pairwise PERMANOVA tests for each pair of groups
  for (i in 1:(nrow(p_value_matrix)-1)) {
    for (j in (i+1):ncol(p_value_matrix)) {
      # Subset the data for each pair of groups
      # group_1 <- pca_data_3d[pca_data_3d$Group == rownames(p_value_matrix)[i], ]
      # group_2 <- pca_data_3d[pca_data_3d$Group == colnames(p_value_matrix)[j], ]
      
       group_1 <- vst_group[vst_group$Group == rownames(p_value_matrix)[i], ]
       group_2 <- vst_group[vst_group$Group == colnames(p_value_matrix)[j], ]
      
      # Combine the two groups into one dataset for distance computation
      combined_data <- rbind(group_1, group_2)
      
      # Compute the distance matrix for the combined data (use Euclidean distance)
      dist_matrix <- dist(combined_data[, -ncol(combined_data)])  # Adjust for more PCs if needed
      
      # Perform PERMANOVA for the pair of groups
      # Run adonis2 with suppressed warnings
       suppressWarnings({
        perm_results <- adonis2(dist_matrix ~ combined_data$Group, permutations = 999)
       })
      
      # Extract the p-value from the 'Pr(>F)' column
      p_value_matrix[i, j] <-  perm_results$`Pr(>F)`[1] # Adjust if column name is different
      p_value_matrix[j, i] <- p_value_matrix[i, j]  # Symmetric matrix
    }
  }
  
  #Extract the upper triangular part of the matrix (excluding the diagonal)
  raw_p_values <- p_value_matrix[upper.tri(p_value_matrix, diag = FALSE)]
  
  # Adjust the p-values using a desired method (e.g., Benjamini-Hochberg)
  adjusted_p_values <- p.adjust(raw_p_values, method = "BH")  # Or "bonferroni", "holm"
  
  # Create a new matrix to store the adjusted p-values
  adjusted_p_value_matrix <- p_value_matrix
  
  # Replace the upper triangular part with the adjusted p-values
  adjusted_p_value_matrix[upper.tri(adjusted_p_value_matrix, diag = FALSE)] <- adjusted_p_values
  
  # Since the matrix is symmetric, copy the upper triangular part to the lower triangular part
  adjusted_p_value_matrix[lower.tri(adjusted_p_value_matrix)] <- t(adjusted_p_value_matrix)[lower.tri(adjusted_p_value_matrix)]
  
  # Print the pairwise p-value matrix adjusted
  print("Pairwise PERMANOVA RESULTS ON Variance Stabilizing Transformation (VST)")
  cat("\n") 
  print("############################")
  print(adjusted_p_value_matrix)
  print("############################")
  cat("\n") 
  }
```

```{r}
# I could have put a simple for loop below but since it was not many and I wanted to seperate datasets in the last HTML I did not
```

# GSE135251

```{r GSE135251, message=FALSE}
GSE135251 <- load_dataset("GSE135251")
GSE135251_meta <- load_meta("GSE135251")
```

```{r GSE135251_3D, message=FALSE}
pca_nas_3d(GSE135251,GSE135251_meta,44,"GSE135251")
```

```{r GSE135251_2D, message=FALSE}
pca_nas_2d(GSE135251,GSE135251_meta,44,"GSE135251")
```

```{r GSE135251_stats}
statistical_report_pca(GSE135251,GSE135251_meta,44,"GSE135251")
```


# GSE193066 

```{r GSE193066}
GSE193066 <- load_dataset("GSE193066")
GSE193066_meta <- load_meta("GSE193066")
```

```{r GSE193066_3D, message=FALSE}
pca_nas_3d(GSE193066,GSE193066_meta,45,"GSE193066")
```

```{r GSE193066_2D, message=FALSE}
pca_nas_2d(GSE193066,GSE193066_meta,45,"GSE193066")
```

```{r GSE193066_stats}
statistical_report_pca(GSE193066,GSE193066_meta,45,"GSE193066")
```

# GSE225740

```{r GSE225740}
GSE225740 <- load_dataset("GSE225740")
GSE225740_meta <- load_meta("GSE225740")
```

```{r GSE225740_3D, message=FALSE}
pca_nas_3d(GSE225740,GSE225740_meta,38,"GSE225740")
```

```{r GSE225740_2D, message=FALSE}
pca_nas_2d(GSE225740,GSE225740_meta,38,"GSE225740")
```

```{r GSE225740_stats}
statistical_report_pca(GSE225740,GSE225740_meta,38,"GSE225740")
```

# GSE130970

```{r GSE130970}
GSE130970 <- load_dataset("GSE130970")
GSE130970_meta <- load_meta("GSE130970")
```

```{r GSE130970_3D, message=FALSE}
pca_nas_3d(GSE130970,GSE130970_meta,46,"GSE130970")
```

```{r GSE130970_2D, message=FALSE}
pca_nas_2d(GSE130970,GSE130970_meta,46,"GSE130970")
```

```{r GSE130970_stats}
statistical_report_pca(GSE130970,GSE130970_meta,46,"GSE130970")
```

# GSE192959

```{r GSE192959}
GSE192959 <- load_dataset("GSE192959")
GSE192959_meta <- load_meta("GSE192959")
```

```{r GSE192959_3D, message=FALSE}
pca_nas_3d(GSE192959,GSE192959_meta,41,"GSE192959")
```

```{r GSE192959_2D, message=FALSE}
pca_nas_2d(GSE192959,GSE192959_meta,41,"GSE192959")
```

```{r GSE192959_stats}
statistical_report_pca(GSE192959,GSE192959_meta,41,"GSE192959")
```

# GSE207310

```{r GSE207310}
GSE207310 <- load_dataset("GSE207310")
GSE207310_meta <- load_meta("GSE207310")
```

```{r GSE207310_3D, message=FALSE}
pca_nas_3d(GSE207310,GSE207310_meta,46,"GSE207310")
```

```{r GSE207310_2D, message=FALSE}
pca_nas_2d(GSE207310,GSE207310_meta,46,"GSE207310")
```

```{r GSE207310_stats}
statistical_report_pca(GSE207310,GSE207310_meta,46,"GSE207310")
```

# GSE162694

```{r GSE162694}
GSE162694 <- load_dataset("GSE162694")
GSE162694_meta <- load_meta("GSE162694")
```

```{r GSE162694_3D, message=FALSE}
pca_nas_3d(GSE162694,GSE162694_meta,44,"GSE162694")
```

```{r GSE162694_2D, message=FALSE}
pca_nas_2d(GSE162694,GSE162694_meta,44,"GSE162694")
```

```{r GSE162694_stats}
statistical_report_pca(GSE162694,GSE162694_meta,44,"GSE162694")
```

# GSE174478

```{r GSE174478}
GSE174478 <- load_dataset("GSE174478")
GSE174478_meta <- load_meta("GSE174478")
```

```{r GSE174478_3D, message=FALSE}
pca_nas_3d(GSE174478,GSE174478_meta,42,"GSE174478")
```

```{r GSE174478_2D, message=FALSE}
pca_nas_2d(GSE174478,GSE174478_meta,42,"GSE174478")
```

```{r GSE174478_stats}
statistical_report_pca(GSE174478,GSE174478_meta,42,"GSE174478")
```

# GSE185051

```{r GSE185051}
GSE185051 <- load_dataset("GSE185051")
GSE185051_meta <- load_meta("GSE185051")
```

```{r GSE185051_3D, message=FALSE}
pca_nas_3d(GSE185051,GSE185051_meta,46,"GSE185051")
```

```{r GSE185051_2D, message=FALSE}
pca_nas_2d(GSE185051,GSE185051_meta,46,"GSE185051")
```

```{r GSE185051_stats}
statistical_report_pca(GSE185051,GSE185051_meta,46,"GSE185051")
```
